---
title: "jointaids"
author: '230195947'
date: "2026-09-11"
output: pdf_document
---

```{r, setup, warning = FALSE}
#install.packages(c("JMbayes2", "nlme", "survival", "GLMMadaptive", "survminer", "knitr", "gtsummary"))
library(JMbayes2)
library(nlme)
library(survival)
library(GLMMadaptive)
library(survminer)
library(tidyverse)
library(knitr)
library(gtsummary)
library(conflicted)
library(flexsurv)
library(broom)
conflicted::conflicts_prefer(JMbayes2::coxph)
conflicts_prefer(JMbayes2::mixed_model)
```

```{r}
data(aids)
data(aids.id)
```

```{r}
km_fit2 <- survfit(Surv(Time, death) ~ drug, data = aids.id)

km_fit2

ggsurvplot(km_fit2, 
           data = aids.id,
           pval = TRUE,             # Adds the log-rank test p-value to the plot
           risk.table = TRUE,       # Adds a table showing patients left over time
           conf.int = TRUE,
           legend.title = "Treatment",
           legend.labs = c("ddC", "ddl"),
           xlab = "Time in Days",
           title = "Kaplan-Meier Survival Estimate for AIDS")
```
The CD4 columns in the JMbayes2 aids datasets refer to the square root of the raw CD4 counts, the purpose of this transformation is to fit an identity link Gaussian mixed effects model to these transformed counts.

To understand the reasoning behind this we must first consider the following. Let $X$ denote the raw CD4 count and $Y = \sqrt{X}$ denote the transformed response. The Gaussian mixed-effects model yields the conditional expectation $\mu_Y = \mathbb{E}[Y]$ and residual variance $\sigma_Y^2$. 

Due to Jensen's Inequality, simply squaring the model's predictions systematically underestimates the true expected count: 
$$ (\mathbb{E}[Y])^2 \leq \mathbb{E}[Y^2] = \mathbb{E}[X] $$

To recover the exact expected value on the raw count scale, we apply a second-order Taylor series expansion to the function $f(Y) = Y^2$ centered around $\mu_Y$:
$$
\begin{align*}
    f(Y) &\approx f(\mu_Y) + f'(\mu_Y)(Y - \mu_Y) + \frac{f''(\mu_Y)}{2}(Y - \mu_Y)^2 + \frac{f'''(\mu_Y)}{6}(Y - \mu_Y)^3 + \dots \\
    Y^2 &\approx \mu_Y^2 + 2\mu_Y(Y - \mu_Y) + \frac{2}{2}(Y - \mu_Y)^2 + \frac{0}{6}(Y - \mu_Y)^3 + \dots
\end{align*}
$$
Third-order or higher Taylor series expansion terms are zero as $\frac{d^3}{dY^3}(Y^2)=0$, therefore we have a strict equality.
Taking the expectation of both sides, the linear term cancels to zero, leaving the exact algebraic correction
$$
\begin{align*}
    \mathbb{E}[Y^2] &= \mathbb{E}[\mu_Y^2] + \mathbb{E}[2\mu_Y(Y - \mu_Y)] + \mathbb{E}[(Y - \mu_Y)^2] \\
    \mathbb{E}[X] &= \mu_Y^2 + 2\mu_Y(\mathbb{E}[Y] - \mathbb{E}[\mu_Y]) + \sigma_Y^2 \\
        \mathbb{E}[X] &= \mu_Y^2 + 2\mu_Y(\mu_Y - \mu_Y) + \sigma_Y^2 \\
    \mathbb{E}[X] &= \mu_Y^2 + \sigma_Y^2
\end{align*}
$$
Thus, the predicted raw counts from the Gaussian model are obtained by adding the model's residual variance to the squared fitted values. This can then be used to calculate the RMSE in the original count scale from the square root scale predicted values generated by the Gaussian mixed effects model. This is then compared directly to the Poisson and negative-binomial model RMSEs. We can also use this property to forecast the CD4 trajectories in their original scale. 

There are several pros and cons to this arrangement.
## Pros
The identity-link Gaussian model simplifies the log-posterior greatly, within the log-posterior the random effects are present in their log-priors and also in the log-likelihood, within the log-likelihood they are found in the context of the inverse-link function. For a log-link this would mean nested in an exponential term, this means that the log-posterior will have a highly irregular shape which will result in the Hamiltonian Monte-Carlo taking longer to converge on the parameters with the highest probability density.

Additionally we have successfully solved for the Jensen's inequality issue. We can fit this model and back-transform the predictions to the original scale without any approximations.

## Cons
We could overcome the Jensen's inequality issue by modelling the count data directly, in particular the negative-binomial distribution will be able to handle discrete counts while also addressing the heavy tail directly with a parameter that gets fitted to account for the overdispersion in the data.

Another issue with simply square-rooting the CD4 counts is that if this transformation doesnt perfectly normalise the data and remove the heteroskedacity, the Gaussian distribution is simply mispecified to the data and there is no more that you can do other than fitting a different exponential family distribution model.

In theory the Gaussian distribution is entirely mispecified to the square-root CD4 counts as the Gaussian distribution has a support of $(-\infty,\infty)$ which not only violates the range of the square-root function, but it also violates the fact that CD4 counts are restricted to positive values as they are a physical quantity. The fitted model will almost certainly assign a negligible predicted probability of a patient having a negative square-root CD4 count, however the non-zero espect implies that this distribution is simply incorrect.

# Alternative Proposal
There is another exponential family distribution that might have worked well for the square rooted counts, however there is one glaring issue. The Gamma distribution has a support of $(0,\infty)$ meaning we lose the unrealistic assumption of negative CD4 counts and we also have an additional dispersion parameter to account for any residual heteroskadicity. The main issue is that the gamma distribution is not defined at $0$ therefore the mass of counts at $0$ cannot be handled without introducing an artificial offset.

This is a massive barrier and as such I will not attempt to use it, as it would require sensitivity analysis of the model to the offset introduced to see how robust the model actually is.

# Survival model 

We can infer which survival model would fit best to this data by first fitting a generalised gamma distribution to the survival times.
$$
\begin{align*}
    \ln(T) &= \mu + \sigma w \\
    w &= \frac{\ln(t) - \mu}{\sigma} \nonumber \\
    f(t \mid \mu, \sigma, Q) &= \frac{|Q| (Q^{-2})^{Q^{-2}}}{\sigma t \, \Gamma(Q^{-2})} \exp\left[ Q^{-2} \left( Qw - \exp(Qw) \right) \right]
\end{align*}
$$

If we set $Q=1$ then we get the following.
$$
\begin{align*}
f(t \mid \mu, \sigma, 1) &= \frac{|1| (1^{-2})^{1^{-2}}}{\sigma t \, \Gamma(1^{-2})} \exp\left[ 1^{-2} \left( 1w - \exp(1w) \right) \right] \\
f(t \mid \mu, \sigma) &= \frac{1}{\sigma t} \exp\left[w - \exp(w)  \right] \\
\end{align*}
$$
Using the definition of $w$ we get,

$$
\frac{dw}{dt}=\frac{d}{dt}\left(\frac{\ln(t) - \mu}{\sigma}\right)=\frac{1}{\sigma t}
$$
We can get the pdf of $w$ by using the following identity,

$$
f_T(t)=\left |\frac{dw}{dt}\right|f_W(w) = \frac{1}{\sigma t} f_W(w)
$$
Therefore,
$$
\begin{align}
f_W(w)&=\exp\left[w - \exp(w)\right]\\
\implies w &\sim \text{Min-EV-Gumbel}(0,1)
\end{align}
$$
We want to get from the minimum extreme value Gumbel distribution to the Weibull, we can do this via the following.


$$
e^w=\exp\left(\frac{\ln(t) - \mu}{\sigma}\right)=\left(e^{\ln t-\mu}\right)^{1/\sigma} =t^{1/\sigma}e^{-\mu/\sigma}.
$$
Therefore, 

$$
f_T(t) = \frac{1}{\sigma t} t^{1/\sigma}e^{-\mu/\sigma} e^{-t^{1/\sigma}e^{-\mu/\sigma}}
$$
Now we set $k=1/\sigma$ and $\lambda=e^\mu$,
$$
\begin{align}
f_T(t) &= k\frac{1}{t} t^{k}(e^{-\mu})^k e^{-t^k(e^{-\mu})^k} \\
&=kt^{k-1}\lambda^{-k}e^{-t^k\lambda^{-k}}\\
&=kt^{k-1}\frac{1}{\lambda\lambda^{k-1}}e^{-(t/\lambda)^k} \\
&=\frac{k}{\lambda}\left(\frac{t}{\lambda}\right)^{k-1}e^{-(t/\lambda)^k} \\
\implies T &\sim \text{Weibull}(k,\lambda)\equiv\text{Weibull}(1/\sigma,e^\mu)
\end{align}
$$
If we now set $k=1 \implies 1/\sigma=1 \implies \sigma=1$ then we have the following.

$$
\begin{align}
f_T(t)&=\frac{1}{\lambda}e^{-\frac{t}{\lambda}} \\
\implies T&\sim\text{Exp}(1/\lambda)
\end{align}
$$
Therefore $Q=\sigma=1 \implies T\sim\text{Exp}(1/\lambda)$

Now we can examine what happens when $Q\to0$. First I manipulate $f_T(t)$ to evaluate the limit easier.
$$\begin{align}
f_T(t) &= \frac{|Q| (Q^{-2})^{Q^{-2}}}{\sigma t \, \Gamma(Q^{-2})} \exp\left[ \frac{ Qw - \exp(Qw)}{Q^2} \right]\\
&= \frac{|Q| (Q^{-2})^{Q^{-2}}}{\sigma t \, \Gamma(Q^{-2})} \exp\left[ \frac{ -1+1+Qw - \exp(Qw)}{Q^2} \right] \\
&= \frac{|Q| (Q^{-2})^{Q^{-2}}}{\sigma t \, \Gamma(Q^{-2})}e^{-1/Q^2} \exp\left[ \frac{1+Qw - \exp(Qw)}{Q^2} \right] \\
&= \frac{|Q| (Q^{-2})^{Q^{-2}}}{\sigma t \, \Gamma(Q^{-2})e^{Q^{-2}}} \exp\left[ \frac{1+Qw - \exp(Qw)}{Q^2} \right] \\
&= C(Q)e^{A(Q)}
\end{align}
$$
Now we take $\lim_{Q\to0}f_T(t)$, to find this limit we need to find $\lim_{Q\to0}C(Q)$ and $\lim_{Q\to0}A(Q)$.

## $\lim_{Q\to0}C(Q)$:
Let $k=Q^{-2}$ therefore $C(k)=\frac{k^k\left|1/\sqrt{k}\right|}{\sigma t \Gamma(k)e^k}=\frac{k^{k-1/2}}{\sigma t\Gamma(k)e^k}$

We know that as $Q\to0$, $k=1/Q^2\to\infty$. Stirling's approximation is an asymptotic relationship for large values of the gamma function, since $k\to\infty$ it becomes an equivalence.

Stirling's approximation tells us that,
$$
\lim_{k\to\infty}\Gamma(k)=\lim_{k\to\infty}\sqrt{\frac{2\pi}{k}}\left(\frac{k}{e}\right)^k = \lim_{k\to\infty}\sqrt{2\pi}k^{k-1/2}e^{-k}
$$

We can now plug this in to demonstrate the following,

$$
\lim_{Q\to0}C(Q)=\lim_{k\to\infty}C(k)=\lim_{k\to\infty}\frac{k^{k-1/2}}{\sigma t\Gamma(k)e^k}=\lim_{k\to\infty}\frac{k^{k-1/2}}{\sigma t\sqrt{2\pi}k^{k-1/2}e^{-k}e^k}=\lim_{k\to\infty}\frac{1}{\sigma t\sqrt{2\pi}}=\frac{1}{\sigma t\sqrt{2\pi}}
$$
## $\lim_{Q\to0}A(Q)$:
By substituting $Q=0$ into $A(Q)$ we get the following $\frac{1+0w - \exp(0w)}{0^2}=\frac{0}{0}$ which is an indeterminate form therefore we can use L'H\^opitals rule to evaluate the limit.

$$
\lim_{Q\to0}A(Q) = \lim_{Q\to0}\left(\frac{\frac{d}{dQ}\left(1+Qw - e^{Qw}\right)}{\frac{d}{dQ}(Q^2)}\right)=\lim_{Q\to0}\left(\frac{w-we^{Qw}}{2Q}\right)=\frac{0}{0}
$$
This again is in an indeterminate form therefore we apply L'H\^opitals rule again to get the following,

$$
\lim_{Q\to0}A(Q) = \lim_{Q\to0}\left(\frac{\frac{d}{dQ}\left(w-we^{Qw}\right)}{\frac{d}{dQ}(2Q)}\right)=\lim_{Q\to0}\left(\frac{-w^2e^{Qw}}{2}\right)=-w^2/2
$$
We have that $\lim_{Q\to0}C(Q)$ and $\lim_{Q\to0}A(Q)$ exist therefore,

$$
\begin{align}
\lim_{Q\to0}f_T(t)&=\lim_{Q\to0}C(Q)e^{A(Q)}\\
&=\left[\lim_{Q\to0}C(Q)\right]e^{[\lim_{Q\to0}A(Q)]}\\
&=\frac{1}{\sigma t\sqrt{2\pi}}e^{-w^2/2}=\left |\frac{dw}{dt}\right|f_W(w) = \frac{1}{\sigma t} f_W(w)
\end{align}
$$
This means that  we have that $f_W(w)=\frac{1}{\sqrt{2\pi}}e^{-w^2/2}\implies w\sim N(0,1)$ therefore from $\ln(T) = \mu + \sigma w$ we can see that $\ln(T)\sim N(\mu,\sigma^2)$ meaning that $T\sim \text{Log-Normal}(\mu,\sigma^2)$

There is one more model nested within the generalised-gamma framework, when we choose $Q=\sigma$ we get the following decomposition


```{r}
generalised_gamma_ph <- flexsurvreg(Surv(Time, death) ~ drug + gender + CD4, 
                          data = aids.id, 
                          dist = "gengamma")
tidy(generalised_gamma_ph)
```

```{r}
aids %>%
  select(CD4) %>%
  pivot_longer(everything()) %>%
  ggplot(aes(x = value)) +
  geom_histogram(bins = 30, fill = "steelblue", color = "black") +
  facet_wrap(~name, scales = "free") +
  theme_minimal() +
  labs(title = "Biomarker Distribution")
```

```{r}
aids$CD4_sqr <- round((aids$CD4)**2)

aids %>%
  select(CD4_sqr) %>%
  pivot_longer(everything()) %>%
  ggplot(aes(x = value)) +
  geom_histogram(bins = 30, fill = "steelblue", color = "black") +
  facet_wrap(~name, scales = "free") +
  theme_minimal() +
  labs(title = "Biomarker Distribution")
```



```{r}
negbinom_CD4 <- mixed_model(CD4_sqr ~ obstime + obstime:drug, 
                        random = ~ obstime | patient, 
                        data = aids,
                        family = negative.binomial)
summary(negbinom_CD4)
```

```{r}
poisson_CD4 <- mixed_model(CD4_sqr ~ obstime + obstime:drug, 
                        random = ~ obstime | patient, 
                        data = aids,
                        family = poisson(link = log))
summary(poisson_CD4)
```

```{r}
cox_aids2 <- coxph(Surv(Time, death) ~ drug + gender, data = aids.id)

summary(cox_aids2)
```

```{r}
joint_fit2 <- jm(cox_aids2, negbinom_CD4, time_var = "obstime", n_iter = 15000, n_burnin = 2500)

summary(joint_fit2)
```

```{r}
# 1. Randomly select exactly 4 patients to match the 2x2 grid layout
set.seed(129) 
random_patients <- sample(aids.id$patient, 4)

# 2. Set up the graphics window for a 2x2 grid with adjusted margins
par(mfrow = c(2, 2), mar = c(4, 4, 3, 1))

# 3. Loop through the 4 patients and generate both longitudinal and survival panels
for (pid in random_patients) {
  
  # Extract this specific patient's chart
  pt_data <- aids[aids$patient == pid, ]
  
  # Forecast their CD4 count trajectory
  pred_long <- predict(joint_fit2, newdata = pt_data, return_newdata = TRUE)
  
  # Forecast their survival/event risk
  pred_event <- predict(joint_fit2, newdata = pt_data, process = "event", return_newdata = TRUE)
  
  # Plot the dual-panel output with explicit Patient ID titles
  plot(pred_long, pred_event, 
       main = paste("Patient ID:", pid),
       xlim = c(0, 25))     
}

# 4. Reset the plotting window back to normal when finished
par(mfrow = c(1, 1), mar = c(5, 4, 4, 2))
```

```{r}
# Calculate the Time-Dependent ROC curve
roc_quiz <- tvROC(joint_fit2, newdata = aids, Tstart = 6, Dt = 6, cores = 1)

# Print the Area Under the Curve (AUC) score
summary(roc_quiz)

# Plot the ROC curve
plot(roc_quiz)
```

```{r}
tvAUC(joint_fit2, newdata = aids, Tstart = 6, Dt = 6, cores = 1)

# Calculate the Brier Score to check prediction accuracy
brier_quiz <- tvBrier(joint_fit2, newdata = aids, Tstart = 6, Dt = 6)

# Print the result
brier_quiz
```

```{r}
set.seed(123)
unique_ids <- unique(aids$patient)
n_folds <- 5

# Create a balanced, randomized vector of fold assignments (1 through 5)
folds <- sample(rep(1:n_folds, length.out = length(unique_ids)))

cv_auc <- numeric(n_folds)
cv_brier <- numeric(n_folds)

# 2. Run the actual refitting loop
for (i in 1:n_folds) {
  cat("Running Fold", i, "of", n_folds, "...\n")
  
  # Identify train and test IDs
  test_ids <- unique_ids[folds == i]
  train_ids <- unique_ids[folds != i]
  
  # Split longitudinal and survival datasets strictly by patient ID
  train_data    <- aids[aids$patient %in% train_ids, ]
  train_data.id <- aids.id[aids.id$patient %in% train_ids, ]
  test_data     <- aids[aids$patient %in% test_ids, ]
  
  # Refit the strictly correct Negative Binomial using Gauss-Kronrod quadrature
  nb_fit <- mixed_model(CD4_sqr ~ obstime + obstime:drug, 
                        random = ~ obstime | patient, 
                        data = train_data,
                        family = negative.binomial)
  
  # Refit the survival submodel
  cox_fit <- coxph(Surv(Time, death) ~ drug + gender, data = train_data.id)
  
  # Refit the joint model
  fold_jm <- jm(cox_fit, nb_fit, time_var = "obstime", n_iter = 8000, n_burnin = 1300)
  summary(fold_jm)
  
  # Test on the remaining 20% unseen data (cores = 1 to prevent socket crashes)
  cv_auc[i]   <- tvAUC(fold_jm, newdata = test_data, Tstart = 6, Dt = 6, cores = 1)$auc
  cv_brier[i] <- tvBrier(fold_jm, newdata = test_data, Tstart = 6, Dt = 6, cores = 1)$Brier
}

# 3. See the true, un-cheated cross-validation scores
cat("\n5-Fold CV AUC:", mean(cv_auc, na.rm = TRUE), "\n")
cat("5-Fold CV Brier:", mean(cv_brier, na.rm = TRUE), "\n")
```